Dynamic Species Interactions in the San Nicolas Island Kelp Forest

### 2018-07-14 EDM_SNI_analysis.Rmd
exploratory analyes at the end

westend.bestE["ymac"] <- 3
westend.bestE["pter"] <- 6

@@

```{r causal vars fxn,message=F,echo=F}
causal_vars <- function(pmat,vars) {
  out <- list()
  for(i in 1:length(vars)) {
    v <- vars[i]
    causes <- colnames(pmat)[pmat[v,]==1]
    causes <- causes[!is.na(causes)]
    out[[v]] <- c(causes,v)
  }
  return(out)
}
westend.causal.vars <- causal_vars(westend.ptot.mat,vars=c("mac","purp","lam","ymac","red","pter"))
```

@@

```{r model perf,echo=F}
# collect model performance statistics
westend.model.perf <- tibble(Species=c("Macrocystis","Purple Urchin","Laminaria","Young Macrocystis","Red Urchin","Pterygophora"),
                             Predictors=purrr::map(westend.causal.vars,.f=function(x){
                               y=namekey$long[match(x,namekey$short)];paste(y,sep=",",collapse=", ")}),
                             Rho = map_dbl(westend.multi.mods,function(x) x$model$rho),
                             MAE=map_dbl(westend.multi.mods,function(x) x$model$mae),
                             mod=c("mac","purp","lam","ymac","red","pter"))
westend.model.perf %>% select(-mod) %>% 
  kable("latex",col.names = c("Modeled Species","Predictors","$\\rho$","MAE"),digits=3,escape=F) %>%
  kable_styling(full_width = F) %>%
  column_spec(1,bold=T)%>%
  column_spec(2, width = "30em")
```

@@

```{r some single plots}
# macrocystis, pterygophora, and temperature
quants<-c(0.05,0.5,0.95)
all.coeffs.with.phys %>%
  filter(type=="algal competition",mod%in% c('mac','pter'),effect %in% c('mac','pter'),external_forcing=="sst")%>%

  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
    geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
    geom_point(col='gray60',alpha=0.6)+
    geom_hline(yintercept = 0,linetype=2,col="white")+
    geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
    scale_color_manual(values=c('mac'="lightblue",'ymac'="gray30",'pter'="red",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),guide="none")+
    labs(x="Normalized SST",y="Interaction Strength")+
    facet_grid(~mod,scales = "free_y")+
  plot_theme_black+
    theme(strip.text=element_blank(),
          plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))

# macrocystis, pterygophora, and NPGO
all.coeffs.with.phys %>%
  filter(type=="algal competition",mod%in% c('mac','pter'),effect %in% c('mac','pter'),external_forcing=="npgo")%>%
  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
  geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+ 
  geom_point(col='gray60',alpha=0.6)+
   geom_hline(yintercept = 0,linetype=2,col="white")+
   geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
   scale_color_manual(values=c('mac'="lightblue",'ymac'="gray30",'pter'="red",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),guide="none")+
   labs(x="Normalized NPGO",y="Interaction Strength")+
   facet_grid(~mod,scales = "free_y")+
  plot_theme_black+
    theme(strip.text=element_blank(),
          plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))

# macrocystis, pterygophora, herbivory, and NPGO
all.coeffs.with.phys %>%
  filter(type=="herbivory",mod%in% c('mac','pter'),effect %in% c('red','purp'),external_forcing=="npgo")%>%
  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
   geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
    geom_point(col='gray60',alpha=0.6)+
   geom_hline(yintercept = 0,linetype=2,col="white")+
   geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
   scale_color_manual(values=c('mac'="lightblue",'ymac'="gray30",'pter'="red",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),guide="none")+
   labs(x="Normalized NPGO",y="Interaction Strength")+
   facet_grid(~mod,scales = "free_y")+
  plot_theme_black+
    theme(strip.text=element_blank(),
          plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))

# macrocystis, pterygophora, and waves
all.coeffs.with.phys %>%
  filter(type=="algal competition",mod%in% c('mac','pter'),effect %in% c('mac','pter'),external_forcing=="waves")%>%
  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
  geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
  geom_point(col='gray60',alpha=0.6)+
   geom_hline(yintercept = 0,linetype=2,col="white")+
   geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
   scale_color_manual(values=c('mac'="lightblue",'ymac'="gray30",'pter'="red",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),guide="none")+
   labs(x="Normalized SWH",y="Interaction Strength")+
   facet_grid(~mod,scales = "free_y")+
  plot_theme_black+
    theme(strip.text=element_blank(),
          plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))

@@

* *Macrocystis pyrifera*
    + Because we have two variables measuring the density of different *Macrocystis* life stages, we can parse out the relative effects of physical conditions on adult vs. juvenile giant kelp.
    + Environmental stress seems to strengthen the interactions between *Laminaria* and *Pterygophora* and *Macrocystis* recruits. Under conditions of elevated MEI, SST, or PDO, or decreased NPGO, the positive effect of *Pterygophora* and the negative effect of *Laminaria* on *Macrocystis* sporophytes are both enhanced. Similar is true for the adult *Macrocystis*: interactions between the giant kelp and the other algae species are elevated under nutrient- or temperature-stressed conditions. Increased wave height (physical disturbance) does not seem to have a large effect on the effect of *Laminaria* and *Pterygophora* on *Macrocystis*
    + The effect of purple urchin on both *Macrocystis* adult and juvenile density is stronger than the red urchin effect, and more variable with changing conditions. Purple urchin herbivory as a negative influence on *Macrocystis* is enhanced with elevated disturbance, MEI, and PDO, and slightly reduced with an increase in the NPGO.
    + The strenghts of intraspecies effects (effects of *Macrocystis* adults on juveniles and vice versa) are also influenced by physical conditions, but perhaps in a different way than might be expected. The positive effect of juvenile *Macrocystis* on adult *Macrocystis* seems to be **enhanced** under nutrient-stressed conditions and higher temperatures, and reduced with greater nutrient availability and increased disturbance.

@@

```{r conditions and interaction strength}
# disturbance. top quartile vs. bottom quartile
all.coeffs.with.phys %>%
  mutate(av_value=abs(value))%>%
  filter(external_forcing=='pdo', type=='algal competition', ext_value>=quantile(westend$pdo,0.75,na.rm=T)|ext_value<=quantile(westend$pdo,0.25,na.rm=T))%>%
  mutate(stress_level=ifelse(ext_value>=quantile(westend$pdo,0.75,na.rm=T),'high','low'))%>%
  ggplot(aes(av_value,..density..,fill=stress_level))+
  geom_density(alpha=0.6)+
  geom_vline(xintercept = 0,linetype=2,color='black')+
  labs(x='Interaction Strength',y='Probability Density')

@@

>>>>

# westend.bestE["ymac"] <- 3
# westend.bestE["pter"] <- 6

@@

```{r causal vars fxn,message=F,echo=F}
# Pull out causal vars from CCM for each species

westend.causal.vars <- purrr::map(c("mac","purp","lam","ymac","red","pter"), function(x) {
  westend.xmap_rast %>%
    filter(predictee==x,!is.na(rho))->out
  out <- c(x,out$predictor)
  out
})

names(westend.causal.vars)=c("mac","purp","lam","ymac","red","pter")

### NOT RUN: VERSION OF MODELS WHERE MULTIVARIATE S-MAPS EQUALS BEST EMBEDDING DIMENSION
# For each, take only the top n causal vars, where n is equal to the best embedding dimension from above
# If n > total number of significant variables from CCM, substitute lagged versions of the variable for dimensions
# westend.causal.vars <- purrr::map(c("mac","purp","lam","ymac","red","pter"),function(x){
#   # number of predictor variables (minus 1 because we use the species' own time series)
#   nvars <- westend.bestE[[x]]
#   westend.xmap_rast %>%
#     select(predictee,predictor,rho)%>%
#     filter(predictee==x)%>%
#     top_n(nvars-1,wt=rho) -> out
#   out <- c(x,out$predictor)
#   if(length(out)<nvars) {
#     ndiff <- nvars-length(out)
#     out <- c(out,paste0(x,1:ndiff))
#   }
#   out
# })
# names(westend.causal.vars) <- c("mac","purp","lam","ymac","red","pter")
# Have to add lagged variables for ymac
# westend <- westend %>% group_by(site) %>% mutate(ymac1=lag(ymac,1)) %>% ungroup()

@@

```{r model perf,echo=F}
# collect model performance statistics
westend.model.perf <- tibble(Species=c("Macrocystis","Purple Urchin","Laminaria","Young Macrocystis","Red Urchin","Pterygophora"),
                             Predictors=purrr::map(westend.causal.vars,.f=function(x){
                               y=namekey$long[match(x,namekey$short)];paste(y,sep=",",collapse=", ")}),
                             Theta = map_dbl(westend.multi.mods,function(x) x$model$theta),
                             Rho = map_dbl(westend.multi.mods,function(x) x$model$rho),
                             MAE=map_dbl(westend.multi.mods,function(x) x$model$mae),
                             mod=c("mac","purp","lam","ymac","red","pter"))
westend.model.perf %>% select(-mod) %>% 
  kable("latex",col.names = c("Modeled Species","Predictors","$\\theta$","$\\rho$","MAE"),digits=3,escape=F) %>%
  kable_styling(full_width = F) %>%
  column_spec(1,bold=T)%>%
  column_spec(2, width = "30em")
```

@@

```{r some single plots}
# macrocystis, pterygophora, and temperature
quants<-c(0.05,0.5,0.95)
all.coeffs.with.phys %>%
  filter(type=="algal competition",mod%in% c('ymac','pter'),effect %in% c('mac','pter'),external_forcing=="sst")%>%

  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
    geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
    geom_point(col='black',alpha=0.6)+
    geom_hline(yintercept = 0,linetype=2,col="black")+
    geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
    scale_color_manual(values=c('mac'="lightblue",'ymac'="gray30",'pter'="red",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),guide="none")+
    labs(x="Normalized SST",y="Interaction Strength")+
    facet_grid(~mod,scales = "free_y")+
    theme(plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))

# macrocystis, pterygophora, and NPGO
all.coeffs.with.phys %>%
  filter(type=="algal competition",mod%in% c('mac','pter'),effect %in% c('mac','pter'),external_forcing=="npgo")%>%
  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
  geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+ 
  geom_point(col='black',alpha=0.6)+
   geom_hline(yintercept = 0,linetype=2,col="black")+
   geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
   scale_color_manual(values=c('mac'="lightblue",'ymac'="gray30",'pter'="red",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),guide="none")+
   labs(x="Normalized NPGO",y="Interaction Strength")+
   facet_grid(~mod,scales = "free_y")+
  theme(plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))

# macrocystis, pterygophora, herbivory, and NPGO
all.coeffs.with.phys %>%
  filter(type=="herbivory",mod%in% c('mac','pter'),effect %in% c('red','purp'),external_forcing=="npgo")%>%
  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
   geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
    geom_point(col='black',alpha=0.6)+
   geom_hline(yintercept = 0,linetype=2,col="black")+
   geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
   scale_color_manual(values=c('mac'="lightblue",'ymac'="gray30",'pter'="red",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),guide="none")+
   labs(x="Normalized NPGO",y="Interaction Strength")+
   facet_grid(~mod,scales = "free_y")+
    theme(plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))

# macrocystis, pterygophora, and waves
all.coeffs.with.phys %>%
  filter(type=="algal competition",mod%in% c('mac','pter'),effect %in% c('mac','pter'),external_forcing=="waves")%>%
  ggplot(aes(ext_value,value,col=factor(effect,levels=c("mac","ymac","pter","lam","purp","red"))))+
  geom_linerange(aes(ymin=lower,ymax=upper),col='gray60')+
  geom_point(col='black',alpha=0.6)+
   geom_hline(yintercept = 0,linetype=2,col="black")+
   geom_quantile(quantiles=quants,lineend = "round",size=1,formula=y~x)+
   scale_color_manual(values=c('mac'="lightblue",'ymac'="gray30",'pter'="red",'lam'="darkcyan",'purp'="mediumpurple4",'red'="darkred"),guide="none")+
   labs(x="Normalized SWH",y="Interaction Strength")+
   facet_grid(~mod,scales = "free_y")+
    theme(strip.text=element_blank(),
          plot.margin=unit(c(0.5, 0.5, 0.5, 0.5), units="line"))

@@

* *Macrocystis pyrifera*
    + Because we have two variables measuring the density of different *Macrocystis* life stages, we can parse out the relative effects of physical conditions on adult vs. juvenile giant kelp.
    + Environmental stress seems to strengthen the interactions between *Laminaria* and *Pterygophora* and *Macrocystis* recruits. Under conditions of elevated MEI, SST, or PDO, or decreased NPGO, the positive effect of *Pterygophora* and the negative effect of *Laminaria* on *Macrocystis* sporophytes are both enhanced. Similar is true for the adult *Macrocystis*: interactions between the giant kelp and the other algae species are elevated under nutrient- or temperature-stressed conditions. Increased wave height (physical disturbance) does not seem to have a large effect on the effect of *Laminaria* and *Pterygophora* on *Macrocystis*
    + The effect of purple urchin on both *Macrocystis* adult and juvenile density is stronger than the red urchin effect, and more variable with changing conditions. Purple urchin herbivory as a negative influence on *Macrocystis* is enhanced with elevated disturbance, MEI, and PDO, and slightly reduced with an increase in the NPGO.
    + The strengths of intraspecies effects (effects of *Macrocystis* adults on juveniles and vice versa) are also influenced by physical conditions, but perhaps in a different way than might be expected. The positive effect of juvenile *Macrocystis* on adult *Macrocystis* seems to be **enhanced** under nutrient-stressed conditions and higher temperatures, and reduced with greater nutrient availability and increased disturbance.

@@

```{r conditions and interaction strength}
# disturbance. top quartile vs. bottom quartile
all.coeffs.with.phys %>%
  mutate(av_value=abs(value))%>%
  filter(external_forcing=='waves', type=='algal competition', ext_value>=quantile(westend$pdo,0.75,na.rm=T)|ext_value<=quantile(westend$pdo,0.25,na.rm=T))%>%
  mutate(stress_level=ifelse(ext_value>=quantile(westend$pdo,0.75,na.rm=T),'high','low'))%>%
  ggplot(aes(log(value),..density..,fill=stress_level))+
  geom_density(alpha=0.6)+
  geom_vline(xintercept = 0,linetype=2,color='black')+
  labs(x='Interaction Strength',y='Probability Density')

@@

### Mean vs. variance in interactions

```{r mean vs variance}
int_mean_sd <-spp.only.mod.coeffs %>%
  filter(!is.na(type))%>%
  mutate(neutral=ifelse(lower<0 & upper>0,1,0))%>%
  group_by(mod,effect,type) %>%
  summarise(mean_int=mean(value,na.rm=T),abs_mean=abs(mean_int),sd_int=sd(value,na.rm=T),n=n(),cv_int=sd_int/abs_mean*100,
            perc_neutral=sum(neutral)/n()*100)%>%
  ungroup()

# Mean interaction strength by interaction type
int_mean_sd %>%
  filter(mod != "ymac",effect != 'ymac')%>%
  group_by(type)%>%
  summarise(mean_str=mean(abs_mean),ci95=sd(abs_mean)/sqrt(n())*1.96)%>%
  ungroup()%>%
  ggplot(aes(type,mean_str))+
  geom_col()+
  geom_errorbar(aes(ymax=mean_str+ci95,ymin=mean_str-ci95),position='dodge',width=0.2)+
  labs(x='Interaction Type',y='Interaction Strength')

# scatterplot mean vs. sd (removing intraspecies interactions)
int_mean_sd %>%
  filter(mod != "ymac",effect != 'ymac')%>%
  ggplot(aes(abs_mean,sd_int))+
  geom_point(aes(color=type),size=5)+
  scale_color_manual(values=c("darkorange1","orangered2","darkorchid1","palegreen3","skyblue1"),name="")+
  geom_smooth(method="lm",se=F,col="black")+
  labs(x="Mean Interaction Strength",y="SD Interaction Strength")

# CV calculated using the correction for sample size (cv*=1+1/4n * cv)
int_mean_sd %>%
  filter(mod != 'ymac',effect!='ymac')%>%
  ggplot(aes(type,cv_int))+
  geom_col()+

  labs(x="Interaction Type",y="CV of Measured Interactions")

  # summarise(mean_cv=mean(cv_int),sd_cv=sd(cv_int),upper=mean_cv+sd_cv/sqrt(n())*1.96,lower=mean_cv-sd_cv/sqrt(n())*1.96)%>%
# geom_errorbar(aes(ymax=upper,ymin=lower),position="dodge",width=0.25)+
```

### Bidirectional interactions (is the effect of A on B the same as B on A)
```{r bidirectional}
int_mean_sd %>%
  mutate(rev_effect=mod,rev_mod=effect) %>%
  left_join(int_mean_sd,by=c("rev_effect"="effect","rev_mod"="mod"))%>%
  mutate(upper1=mean_int.x+sd_int.x,lower1=mean_int.x-sd_int.x,upper2=mean_int.y+sd_int.y,lower2=mean_int.y-sd_int.y)-> int_mean_sd2

int_mean_sd2 %>%
  filter(type.x %in% c('algal competition','urchin competition'),type.y %in% c('algal competition','urchin competition'))%>%
  ggplot(aes(mean_int.x,mean_int.y))+
  geom_point()+
  geom_smooth(method="lm")+
  geom_abline(slope=1,intercept=0,linetype=2)+
  geom_errorbar(aes(ymin=lower2,ymax=upper2))+
  geom_errorbarh(aes(xmin=lower1,xmax=upper1))+
  labs(x='Effect of A on B',y='Effect of B on A')
```

### Within-model cross correlation of effects

```{r cross correlation of effect}
cc_effects <- purrr::map(c("mac","purp","lam","ymac","red","pter"),function(x){
  spp.only.mod.coeffs %>% filter(mod==x,effect!='const') %>%
    select(time,period,effect,value)%>%spread(effect,value) %>%
    left_join(select(westend,one_of(phys.vars),period),by="period")%>%distinct()%>%
    select(-time,-period)%>%
    cor(use="complete.obs")
})
names(cc_effects) <- c("mac","purp","lam","ymac","red","pter")
cc_effects_plot <- function(model,effect1,effect2){
  spp.only.mod.coeffs %>% filter(mod==model,effect%in%c(effect1,effect2)) %>%
    select(time,period,effect,value)%>%spread(effect,value)%>%
    left_join(select(westend,one_of(phys.vars),period),by="period")%>%distinct()%>%
    ggplot(aes_string(effect1,effect2))+geom_point()+geom_smooth(method='lm',se=F)
}
```
